DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models

### 2017-08-13 DCMIP2016-Part1-Abstract.tex
Updates based on returned reviews

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package which can be downloaded from the different journal webpages.
%% For further assistance please contact the Copernicus Publications at: publications@copernicus.org
%% http://publications.copernicus.org

%% Please use the following documentclass and Journal Abbreviations for Discussion Papers and Final Revised Papers.

%% 2-Column Papers and Discussion Papers
\documentclass[gmd, manuscript]{copernicus}

%% Journal Abbreviations (Please use the same for Discussion Papers and Final Revised Papers)

% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Advances in Geosciences (adgeo)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% ASTRA Proceedings (ap)
% Atmospheric Measurement Techniques (amt)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Biogeosciences (bg)
% Climate of the Past (cp)
% Drinking Water Engineering and Science (dwes)
% Earth System Dynamics (esd)
% Earth Surface Dynamics (esurf)
% Earth System Science Data (essd)
% Fossil Record (fr)
% Geographica Helvetica (gh)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% Geothermal Energy Science (gtes)
% Hydrology and Earth System Sciences (hess)
% History of Geo- and Space Sciences (hgss)
% Journal of Sensors and Sensor Systems (jsss)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Primate Biology (pb)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

\usepackage{subcaption}

% Custom commands
\newcommand{\vb}{\mathbf}
\newcommand{\vg}{\boldsymbol}
\newcommand{\mat}{\mathsf}
\newcommand{\diff}[2]{\frac{d #1}{d #2}}
\newcommand{\diffsq}[2]{\frac{d^2 #1}{{d #2}^2}}
\newcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\pdiffsq}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}

% User-defined mathematical symbols
\renewcommand{\emph}[1]{{\color{red}\textbf{#1}}}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\p}[1]{\partial #1}
%Then, writing a partial is as simple as $\pd{J}{x}$.
\newcommand{\ol}[1]{{\overline #1}}
\newcommand{\mbf}[1]{{\mathbf #1}}
\newcommand{\mbfol}[1]{\overline{{\mathbf #1}}}
\newcommand{\wdh}[1]{{\widehat #1}}
\newcommand{\wdt}[1]{{\widetilde #1}}
\newcommand{\sss}[1]{{\scriptscriptstyle #1}}
\newcommand{\scs}[1]{{\scriptstyle #1}}

@@

\title{DCMIP2016:  A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models}

% \Author[affil]{given_name}{surname}

\Author[1]{Paul A.}{Ullrich}
\Author[2]{Christiane}{Jablonowski}
\Author[3]{James}{Kent}
\Author[4]{Peter H.}{Lauritzen}
\Author[4]{Ramachandran}{Nair}
\Author[5]{Kevin A.}{Reed}
\Author[4]{Colin M.}{Zarzycki}

\Author[6]{David M.}{Hall}

\Author[7]{Don}{Dazlich}
\Author[7]{Ross}{Heikes}
\Author[7]{Celal}{Konor}
\Author[7]{David}{Randall}

\Author[8]{Thomas}{Dubos}
\Author[8]{Yann}{Meurdesoif}

\Author[9]{Xi}{Chen}
\Author[9]{Lucas}{Harris}

\Author[10]{Christian}{K\"uhnlein}

\Author[11]{Vivian}{Lee}
\Author[11]{Abdessamad}{Qaddouri}
\Author[11]{Claude}{Girard}

\Author[12]{Marco}{Giorgetta}
\Author[13]{Daniel}{Reinert}

\Author[4]{Joseph}{Klemp}
\Author[4]{Sang-Hun}{Park}
\Author[4]{William}{Skamarock}

\Author[14]{Hiroaki}{Miura}
\Author[14]{Tomoki}{Ohno}
\Author[15]{Ryuji}{Yoshida}

\Author[16]{Robert}{Walko}

\Author[17]{Alex}{Reinecke}
\Author[17]{Kevin}{Viner}

%\Author[18]{Elijah}{Goodfriend}
%\Author[18]{Hans}{Johansen}

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{University of Colorado, Boulder}
\affil[7]{Colorado State University}
\affil[8]{Institut Pierre-Simon Laplace (IPSL)}
\affil[9]{Geophysical Fluid Dynamics Laboratory (GFDL)}
\affil[10]{European Centre for Medium-Range Weather Forecasts (ECMWF)}
\affil[11]{Environment and Climate Change Canada}
\affil[12]{Max Planck Institute for Meteorology}
\affil[13]{Deutscher Wetterdienst (DWD)}
\affil[14]{University of Tokyo}
\affil[15]{RIKEN}
\affil[16]{University of Miami}
\affil[17]{Naval Research Laboratory}

%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

\runningtitle{DCMIP2016: A Review of Non-hydrostatic Dynamical Core Design and Intercomparison}

\runningauthor{Ullrich, et al.}

\correspondence{Paul A. Ullrich (paullrich@ucdavis.edu)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.  These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.  In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.  To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.  This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
\end{abstract}

\introduction  %% \introduction[modified heading if necessary]

The Dynamical Core Model Intercomparison Project (DCMIP) is an ongoing effort targeting the intercomparison of a fundamental component of global atmospheric modeling systems: the dynamical core.  Although this component's role is simply to solve the equations of fluid motion governing atmospheric dynamics (the Navi\'er-Stokes equations), there are numerous confounding factors and compromises that arise from making global simulations computationally feasible.  These factors include the choice of model grid, variable placement, vertical coordinate, prognostic equations, representation of topography, numerical method, temporal discretization, physics/dynamics coupling frequency, and the manner in which artificial diffusion, stabilization, filters and/or energy/mass fixers are applied.

To advance the intercomparison project and provide a unique educational opportunity for students, DCMIP hosted a multidisciplinary two-week summer school and model intercomparison project, held at the National Center for Atmospheric Research (NCAR) in June 2016, that invited graduate students, postdocs, atmospheric modelers, expert lecturers and computer specialists to create a stimulating, unique and hands-on driven learning environment. The 2016 workshop and summer school followed from earlier DCMIP and dynamical core workshops (held in 2012 and 2008, respectively), and other model intercomparison efforts.  Its goals were to provide an international forum for discussing outstanding issues in global atmospheric models, and provide a unique training experience for the future generation of climate scientists. Special attention was paid to the role of simplified physical parameterizations, physics-dynamics coupling, non-hydrostatic atmospheric modeling, and variable-resolution global modeling. The summer school and model intercomparison project promoted active learning, innovation, discovery, mentorship and the integration of science and education.  Modeling groups were then invited to contribute model descriptions and results to the intercomparison effort for publication.

The summer school directly benefited its participants by providing a unique educational experience and an opportunity to interact with modeling teams from around the world.  The workshop is expected to have further repercussions on the development of operational atmospheric modeling systems, by allowing modeling groups to assess their models in the context of the global dynamical core ecosystem.  Past and present intercomparison efforts have been leveraged by modeling groups to improve their own models, in turn leading to a positive impact on the quality of weather and climate simulations.  The workshop component of DCMIP has also advanced our knowledge of (1) the relative behaviors exhibited by atmospheric dynamical cores, (2) differences that arise among mechanisms for coupling the physical parameterizations and dynamical core, and (3) the impacts of variable-resolution refinement regions and transition zones in global atmospheric simulations.  Notably, the use of idealized test cases to isolate specific phenomena gave us a unique opportunity to assess specific differences that arise due to the choice of dynamical core.  Another important outcome of the workshop was the development of a standard test case suite and benchmark set of simulations that can be used for assessment of any future dynamical core.  The test cases introduced in the 2016 workshop build on the previous DCMIP test case suites \citep{jablonowski2008idealized, ullrich2012dynamical} with tests that now incorporate simplified moist physics.

This paper is the first in a series of papers documenting the results of this workshop.  Its purpose is twofold:  First, to review the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling; and second, to provide a mechanism to understand the differences that arise in the test cases of later papers in this series.  For ease of reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.  Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.  The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  section \ref{sec:HorizontalDiscretization} describes aspects of the horizontal discretization, including model grids and horizontal placement of prognostic variables; section \ref{sec:VerticalDiscretization} describes the vertical placement of model variables and choice of vertical coordinates; section \ref{sec:PrognosticEquations} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.  The summary and conclusions then follow in section \ref{sec:Conclusions}.  Finally, appendix \ref{sec:EquationSets} provides a comprehensive overview of the various forms the Navi\'er-Stokes equations take in dynamical cores, and has been included as a resource for dynamical core developers.

\begin{table}[p]
\caption{A standard list of symbols used throughout this paper and in the DCMIP.} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$s$ & Vertical model coordinate \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind velocity \\
$v$ & Meridional wind velocity \\
$w$ & Vertical wind velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$\theta_{il}$ & Ice-liquid potential temperature \\
$\theta_\rho$ & Density potential temperature \\
%$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Specific humidity \\
$q_v$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Dynamical cores} \label{sec:DynamicalCores}

This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  In total, simulation results and model descriptions have been submitted from eleven dynamical cores (see Table \ref{tab:Models}).  The prognostic variables employed and horizontal discretizations for these dynamical cores are summarized in Table \ref{tab:ModelsHorizontal}.  The vertical staggering of variables and vertical coordinate choice is summarized in Table \ref{tab:ModelsVertical}.  Standard options for diffusion, stabilization, filters, or fixers along with the temporal discretization for these models is summarized in Table \ref{tab:ModelsTemporalDiffusion}.  A brief description of each participant model follows, focused on the unique features and decisions underlying the model design.

%Although all models solve the Navi\'er-Stokes equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.  The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.

\begin{table}[p]
\caption{Participating modeling centers and associated dynamical cores that have submitted a model description and/or simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Short Name & Long Name & Modeling Center or Group \\ \hline 
ACME--A & Atmosphere model of the Accelerated Climate Model for Energy & Sandia National Laboratories and \\
& & University of Colorado, Boulder, USA \\
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
OLAM & Ocean Land Atmosphere Model & Duke University / University of Miami, USA \\
TEMPEST & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only three numerical methods are represented among participating models, namely finite-difference (FD), finite-volume (FV), and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Short Name & Equation Set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
ACME--A & H/NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta$, $\Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q_i$ & Octahedral (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\dot{\zeta}$, $T_v$, $p$, $q_i$ & Yin-Yang (\S \ref{sec:grid_yinyang}) & FD & C-grid  \\
ICON & NH (D) & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q_i$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q_i$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH (D) & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho \theta_{il}$, $\rho q_i$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
TEMPEST & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q_i$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
ACME--A & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FV$^3$ & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FVM & Co-located & fixed height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
TEMPEST & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
ACME--A & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & 4th-order horizontal hyperviscosity & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagorinsky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
TEMPEST & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

%Do not include information on the physical parameterizations used by the modeling system.  Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}

\subsection{Accelerated Climate Model for Energy--Atmosphere (ACME--A)}

The ACME--A model has much in common with the Community Atmosphere Spectral Element Model (CAM-SE) \citep{dennis2012cam} as both share a common origin in the High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible}. ACME-A employs both a hydrostatic model and an experimental non-hydrostatic compressible shallow-atmosphere model. Both variants are designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. ACME-A partitions the globe horizontally via an unstructured grid of quadrilateral elements arranged in a cubed-sphere configuration, although unstructured regionally-refined meshes with conforming edges may also be employed. The fluid equations are discretized using dimensional splitting, with a nodal 4th-order spectral element discretization in the horizontal and a vertical floating Lagrangian levels in hybrid terrain-following pressure coordinates $\eta$.
Vertical operators are based on the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy}. All fields are co-located in the horizontal, in the sense that they share the same 4th-order basis functions.

%Prognostic variables (u, v, w, $\rho_s$, $\rho_s \theta$, and $\Phi$) are placed on layer midpoints and diagnostic vertical fluxes ($m \dot \eta$) computed on layer interfaces.

%The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME--A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following pressure coordinate in the vertical. The fluid equations are discretized using dimension splitting, with a nodal 4th-order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th-order basis functions. Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.

\subsection{Colorado State University Model (CSU)}

The CSU model uses an optimized geodesic grid to discretize the sphere \citep{heikes1995numerical,heikes2013optimized}, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.  The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature \citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

\subsection{FV Cubed (FV$^3$)}

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid as described by \citet{LR1997QJR} (also see section \ref{sec:HorizontalStaggering}). Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.  Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

The prognostic horizontal winds are stored in the native Gnomonic local coordinate. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.

\subsection{Finite-Volume Module (FVM) of the Integrated Forecasting System}
\label{sec:FVM}

The Finite-Volume Module (FVM) of the Integrated Forecasting System (IFS) is currently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016,kuehnleinJCP2017,smolarkiewiczetalJCP2017}. FVM solves the 
compressible Euler equations in a geospherical framework \citep{szmelter2010,smolarkiewiczetalJCP2016}. 
A centered two-time-level semi-implicit 
integration scheme is employed with 3D implicit treatment of acoustic, buoyant, and rotational modes 
\citep{smolarkiewiczJCP2014}. The associated 3D Helmholtz problem is solved iteratively using a bespoke 
preconditioned Generalised Conjugate Residual approach. The integration procedure uses the 
non-oscillatory MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
The horizontal spatial discretization is fully unstructured finite-volume using the median-dual approach. This is 
combined with a structured-grid finite-difference discretization in the vertical direction. In both the horizontal and 
the vertical discretization all variables are co-located. 
The median-dual finite-volume mesh in the horizontal is developed about the nodes of the octahedral reduced 
Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral reduced Gaussian 
grid is also employed in the spectral-transform dynamical core of the presently operational IFS at ECMWF, which facilitates 
interoperability of the two formulations. However, FVM is not restricted to this grid and offers capabilities towards 
broad classes of meshes including adaptivity \citep[e.g.][]{szmelter2010,kuehnleinJCP2012}.
%No explicit diffusion or filtering is applied in FVM for DCMIP, unless it is required for particular test cases.

%The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
%ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
%equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
%are available by means of simple switches. The formulation incorporates 
%a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
%A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
%of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
%Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
%Conjugate Residual approach. The dynamics time step is adapted 
%at every time step according to a given maximum advective CFL number (typically somewhat smaller 
%than 1). The physics time step is identical to the dynamics time step. 
% 
%The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
%advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
%The horizontal spatial discretization is fully unstructured finite-volume using the 
%median-dual approach. This is combined with a structured-grid finite-difference approach in the 
%vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
%In both the horizontal and the vertical discretization, all prognostic variables are 
%co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
%of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
%reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
%IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
%not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity. 
%
%No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
%diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
%boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
%the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
%the form of a Rayleigh-type forcing to the prognostic variables.

%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.

%Neither explicit diffusion/filters nor decentering of the time-stepping scheme is applied in DCMIP, 
%No explicit diffusion is applied in FVM in DCMIP, apart from the explicit momentum dissipation and 
%scalar diffusion required for some of the test cases.  FVM uses adaptive time stepping for optimal
%efficiency and accuracy.

%FVM employs distributed- and shared-memory parallelisation using the MPI and OpenMP standards, 
%respectively. %The underlying distributed-memory data structures in FVM are provided by the Atlas 
%framework developed at ECMWF.

\subsection{Global Environmental Multiscale (GEM) model} \label{sec:GEM_core}

>>>>

\documentclass[11pt]{article}

@@

\noindent \textbf{DCMIP2016:  A Review of Non-hydrostatic Dynamical Core Design and Intercomparison of Participating Models} \\

\noindent {Paul A. Ullrich, Christiane Jablonowski, James Kent, Peter H. Lauritzen, Ramachandran Nair, Kevin A. Reed, Colin M. Zarzycki, David M. Hall, Don Dazlich, Ross Heikes, Celal Konor, David Randall, Thomas Dubos, Yann Meurdesoif, Xi Chen, Lucas Harris, Christian K\"uhnlein, Vivian Lee, Abdessamad Qaddouri, Claude Girard, Marco Giorgetta, Daniel Reinert, Joseph Klemp, Sang-Hun Park, William Skamarock, Hiroaki Miura, Tomoki Ohno,
Ryuji Yoshida, Robert Walko, Alex Reinecke, and Kevin Viner} \\

\noindent Atmospheric dynamical cores are a fundamental component of global atmospheric modeling systems, and are responsible for capturing the dynamical behavior of the Earth's atmosphere via numerical integration of the Navi\'er-Stokes equations.  These systems have existed in one form or another for over half of a century, with the earliest discretizations having now evolved into a complex ecosystem of algorithms and computational strategies.  In essence, no two dynamical cores are alike, and their individual successes suggest that no perfect model exists.  To better understand modern dynamical cores, this paper aims to provide a comprehensive review of eleven dynamical cores, drawn from modeling centers and groups that participated in the 2016 Dynamical Core Model Intercomparison Project (DCMIP) workshop and summer school.  This review includes choice of model grid, variable placement, vertical coordinate, prognostic equations, temporal discretization, and the diffusion, stabilization, filters and fixers employed by each system.
